Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-04-01 analysis/paper.qmd
Add references

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  Modelled ranges do not reliably predict the observed  occurrence of specific 
  species at archaeological sites, but are aligned with more general trends in
  the relative abundance of species.<!--
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.-->
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

@@

The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@HarlanZohary1966; @ZoharyHopf2000; @Diamond2002].
These species supported uniquely dense and complex Late Epipalaeolithic [15–11.7 ka, @BarYosef1998; @MaherEtAl2012] societies based on foraging [@HillmanEtAl1989; @Colledge2001; @WeissEtAl2004] and eventually plant management and pre-domestication cultivation [@Colledge2001; @WeissEtAl2006; @Harris2007; @WillcoxEtAl2008].
The first agro-ecosystems emerged as these plants were domesticated in the Pre-Pottery Neolithic period (11.7–8.5 ka), shaped by the broader ecosystem in which it was embedded.

@@

The transition to agriculture represents one of the most fundamental changes in human history.
West Asia is one of the regions where this process has been studied in the most detail:
decades of research have traced the gradual development of a Neolithic way of life and the changes that occurred in the plant species and their geographical distribution as a result.
Although archaeobotanical assemblages can be biased due to issues of preservation, sampling, recovery techniques, and lab procedures [Dennel 1976, Hastorf and Popper 1988] —and although they include not just food remains but plant resources that were used for other purposes or arrived at the site accidentally [@HastorfPopper1988]—large-scale studies have still revealed coherent patterns in the exploitation of plants over time [Colledge et al. 2004, Arranz-Otaegui et al. 2016].

The possibility of an abrupt, geographically-constrained process of plant domestication was proposed in the 1990s [@HillmanDavies1990; @HillmanDavies1992; @HeunEtAl1997; @OzkanEtAl2002] and developed as an explanatory model in the 2000s [@LevYadunEtAl2000; @GopherEtAl2001; @AbboEtAl2005].
As part of this model, some authors [Lev-Yadun, Gopher and Abbo 2000, Gopher, Abbo and Lev-Yadun 2001, Abbo, Lev-Yadun and Gopher 2010, 2012] argued that eight plant species, collectively referred to as 'founder crops' or the 'Neolithic crop package' [Zohary and Hopf 1988] were selected and domesticated once, without any phase of pre-domestication cultivation [Abbo, Lev-Yadun and Gopher 2011, p. 177)].
This process could have been rapid under strong artificial selection [Hillman and Davies 1990, 1992] and may have occurred in a single region or 'core area' – generally located in southeast Turkey [Ladizinsky and Adler 1976, Heun et al. 1997, Özkan et al. 2002, 2005, Mori et al. 2003, Luo et al. 2007]. 
From this single point of origin, it was supposed that domesticated or semi-domesticated plants radiated outwards to other regions [Abbo et al. 2006, Kilian et al. 2007, Özkan et al. 2011].

The 'short gestation' paradigm was challenged by others [Helbaek 1969, Harris 1989, Kislev 1989, Colledge 2001, Weiss et al. 2004, Willcox et al. 2008, Fuller et al. 2018]. 
@Helbaek1966 argued that before the appearance of domesticated plants, a phase of cultivation of wild seeds must have taken place. 
The existence of a phase of cultivation of morphologically wild cereals or 'pre-domestication cultivation' was identified in the archaeological record through the study of plant domestication traits such as grain size, shattering v. non-shattering rachises. 
The archaeobotanical evidence shows that during the Pre-Pottery Neolithic A (PPNA) cereals exhibited sizes similar to those recorded in domestic species, but their dispersal mechanism was still the same as the one present in morphologically-wild species [i.e. shattering, see Helbaek 1966, Kislev 1989, Hillman et al. 2000, Colledge 2001, Willcox et al. 2008].
This evidence suggested that wild cereal stands could have been cultivated for as much as a thousand years before non-shattering domestic forms became prevalent in the archaeological record [Tanno and Willcox 2006, 2012, Arranz-Otaegui et al. 2016)].
Additional archaeobotanical [Colledge 2001, Willcox et al. 2008, 2009, Riehl et al. 2013, Arranz-Otaegui et al. 2016, Weide et al. 2018, Douché and Willcox 2018, Whitlam et al. 2018] and genetic data [Badr et al. 2000, Molina-Cano et al. 2005, Kilian et al. 2007, Özkan et al. 2011, Lob et al. 2023 VHA] in recent years has further challenged the short-gestation model to explain the origins of plant domestication and agriculture in West Asia.

Similarly, the concept of a limited set of eight 'founder crops' [@ZoharyHopf1988] that were the first species cultivated, domesticated and then spread as the basis of Neolithic agricultural systems, is not supported by the latest evidence.
Our previous analyses of the composition of available archaeobotanical datasets shows that these crops were of marginal importance during the Epipalaeolithic period [@ArranzOtaeguiEtAl2018] and that Neolithic subsistence did not rely either solely or primarily on the exploitation of these species [Arranz-OtaeguiRoe2023]. 
Instead, multiple species of grasses, legumes, fruits, nuts, and other plants were exploited over the Late Pleistocene-Early Holocene transition in southwest Asia.

@@

The study of the natural distribution of the progenitors of domesticated crops has been a central part of discussions on the origins of agriculture and plant domestication from the beginning.
@Humboldt1807 acknowledged the importance of the natural distribution of wild species to explain the origin and domestication of crops like spelt and rye.
@Candolle1820 integrated the study of plant ecology and biogeography and influenced @Darwin1859, who later reflected in detail about the geographical distribution of plants and species diversity. 
At that time, there was intense debate about whether there were single or multiple "centres of creation" of species. 
Researchers aimed to evaluate whether plant and animal species emerged in the same locations where they were currently distributed.

After Darwin, this early interest in crop origins evolved into more specific discussions about the "centres of plant domestication".
Vavilov [@Vavilov1926; @Vavilov1950] can be considered one of the first authors that sought to determine the number of regions in which plants had been independently domesticated. 
His main method was 'differential phytogeography' [@Vavilov1950]: 
he classified the variation within a crop and established the regions of maximum diversity, to locate the geographic regions in which crops originated. 
Using this method, Vavilov suggested that there were at least eight centres of origin. 
However, his work was later criticised by @Harlan1971, who argued that 'centres of origin' and 'centres of diversity' had to be separated. 
For Harlan, a 'centre' was as an "area in which things originate and out of which things are dispersed" [@Harlan1971, p. 468], and he suggested that three main centres of origin of domesticated crops existed.
He further indicated that Vavilov's approach to the question was simplistic and that more data proxies had to be considered (e.g. archaeology, history, geology…), an approach more in the tradition of @Candolle1909. 
Indeed, the inclusion of archaeobotany and genetics in the last decades, together with the study of wild relative distributions has been fundamental in characterising the origins of agriculture [@FullerColledge2008]. 
As a result of modern interdisciplinary studies, the number of recognised centres of plant domestication has increased considerably, from the three centres suggested by Harlan in 1971 to the six to eight centres argued for in the 1990s [@Smith1995] and up to as much as 24 potential centres reported in 2009 [@PuruggananFuller2009; see also @Fuller2010].

Biogeographic research into the centres of origin and/or domestication of crops has directly informed broader understanding of the process of agricultural origins.
In particular, Childe's *Man Makes Himself* [-@Childe1936] both set the agenda for much of the archaeological work that took place between the end of the Second World War and the 80s [e.g. @BraidwoodBraidwood1950; @Kenyon1952; @Kenyon1960; @Braidwood1981; @HoleFlannery1968; @Moore1975] and stimulated theoretical discussions on the subject [e.g. @Sauer1952; @Harris1969; @Harris1977; @Harris1989; @Flannery1973; @Hawkes1969]. 
Childe (unlike for example @Pumpelly before him) correctly located the centre of origin of European agriculture in West Asia, but not based on the region's prehistoric archaeological record, which at that point was barely known.
Instead he was guided to the region by biogeographic work by Vavilov and @PeakeFleure1927;
only later was this prediction validated by archaeological work on the Epipalaeolithic and Neolithic of Palestine [@Boyd2018].

Contemporary research on crop origins was pioneered by Harlan and Zohary, who compared the current distribution of the wild progenitors of domesticated plants in southwest Asia [@HarlanZohary1966; @Zohary1969; @Zohary1973; @ZoharyHopf1973] to the rapidly-expanding archaeobotanical record [@Harlan1971; @Harlan1977; see also @ZoharyHopf1988; @HarlanZohary1966].
They both were interested in evaluating which were the wild ancestors of domesticated crops and studying their natural distribution to understand their domestication process [Zohary 1969, Zohary 1973, 1992, Zohary and Spiegel-Roy 1975]. 
Indeed, the natural distribution of the wild relatives of domestic plant species was later used as a criterion to infer 'pre-domestic cultivation' in the archaeological record. 
For example, the presence of seeds of chickpea at Jericho [Hopf 1983, see also Hopf 1969] led the author to interpret the remains as cultivars, as the natural distribution of the wild form of chickpea was located further to the north. 
The same rationale was applied to the einkorn remains found at several Pre-Pottery Neolithic sites in the southern Levant [Hopf 1969, Colledge 2001], as the wild progenitors of this species was thought to be restricted to the northern Levantine area [Heun et al. 1997, 2009, Zohary and Hopf 2000]. 
The same idea—presence of plants outside their natural range—has been repeated in the literature more recently by several other authors [Tanno and Willcox 2006, Willcox et al. 2008, Hilman et al. 2000].

Despite the importance of biogeography in the development and validation of hypotheses regarding the origins of agriculture, there have been few studies of the wild range of specific crop progenitors or other relevant plant species [cf. for domestic animals, e.g. @YeomansEtAl2017].
Observations regarding translocation or range expansion must therefore rely on a relatively rough and ahistoric notion of a species' 'natural distribution' – that is, one based primarily on contemporary or recent-historic occurrences.
Yet we know there has been considerable climatic and environmental change in West Asia since the terminal Pleistocene [@JonesEtAl2019], so it is very unlikely that these ranges were in fact static.
Reconstructions of broader environments have attempted to trace their fluctuations through time, either for the entire region [e.g. @VanZeistBottema1991; Hillman2000] or parts of it [e.g. @Cordova2007], but these are at the level of the vegetation zone rather than individual species.
They also invariably rely on what might be called 'expert interpolation' (where the author composes a map based on his or her own knowledge of the relevant data) rather than an explicit modelling process.
This makes it difficult, if not impossible, for users of such reconstructions to understand exactly how they were derived or what could explain, for example, the significant discrepencies between the predictions of different experts.

@@

Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye). 
Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2000; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].

@@

The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations. 
Since they do accurately predict the test dataset <!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
Another obvious flaw in our methodology is that the time slices used for palaeoclimatic reconstruction are very broad—each covering around two millennia—and therefore potentially unrepresentative of the environment around sites at the specific time at which they were occupied.
The variable quality of the archaeological test dataset, especially in terms of chronology, is also a plausible factor.

@@

"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.

* Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
* Discrepancies suggests several areas for further research and methodological development

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  The models performance in predicting the occurrence of specific species at 
  archaeological sites is highly variable, but in aggregate the predictions are
  coherent and align with broad-scale trends in the ubiquity and relative 
  abundance of species in the archaeological record.<!--
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.-->
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

@@

The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@HarlanZohary1966; @Diamond2002; @ZoharyEtAl2012].
These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies [@BarYosef1998; @MaherEtAl2012] based on foraging [@HarrisHillman1989; @Colledge2001; @WeissEtAl2004] and eventually plant management and pre-domestication cultivation [@Colledge2001; @WeissEtAl2006; @Harris2007; @WillcoxEtAl2008].
The first agro-ecosystems emerged as these plants were domesticated in the Pre-Pottery Neolithic period (11.7–8.5 ka) and were shaped by the broader ecosystem in which it was embedded.

@@

The transition to agriculture represents one of the most fundamental changes in human history.
West Asia is one of the regions where this process has been studied in the most detail:
decades of research have traced the gradual development of a Neolithic way of life and the changes that occurred in the plant species and their geographical distribution as a result.
Although archaeobotanical assemblages can be biased due to issues of preservation, sampling, recovery techniques, and lab procedures [@Dennel1976; @HastorfPopper1988]—and although they include not just food remains but plant resources that were used for other purposes or arrived at the site accidentally [@HastorfPopper1988]—large-scale studies have still revealed coherent patterns in the exploitation of plants over time [@ColledgeEtAl2004; @ArranzOtaeguiEtAl2016].

The possibility of an abrupt, geographically-constrained process of plant domestication was proposed in the 1990s [@HillmanDavies1990; @HillmanDavies1992; @HeunEtAl1997; @OzkanEtAl2002] and developed as an explanatory model in the 2000s [@LevYadunEtAl2000; @GopherEtAl2001; @AbboEtAl2005].
As part of this model, some authors [@LevYadunEtAl2000; @GopherEtAl2001; @AbboEtAl2010; @AbboEtAl2012] argued that eight plant species, collectively referred to as 'founder crops' or the 'Neolithic crop package' [@ZoharyHopf1988] were selected and domesticated once, without any phase of pre-domestication cultivation [@AbboEtAl2011, p. 177].
This process could have been rapid under strong artificial selection [@HillmanDavies1990; @HillmanDavies1992] and may have occurred in a single region or 'core area' – generally located in southeast Turkey [@LadizinskyAdler1976; @HeunEtAl1997; @OzkanEtAl2002; @OzkanEtAl2005; @MoriEtAl2003; @LuoEtAl2007]. 
From this single point of origin, it was supposed that domesticated or semi-domesticated plants radiated outwards to other regions [@AbboEtAl2006; @KilianEtAl2007; @OzkanEtAl2011].

The 'short gestation' paradigm was challenged by others [@Helbaek1969; @Harris1989; @Kislev1989; @Colledge2001; @WeissEtAl2004; @WillcoxEtAl2008; @FullerEtAl2018]. 
Helbæk [in @Kirkbride1966] argued that before the appearance of domesticated plants, a phase of cultivation of wild seeds must have taken place. 
The existence of a phase of cultivation of morphologically wild cereals or 'pre-domestication cultivation' was identified in the archaeological record through the study of plant domestication traits such as grain size, shattering v. non-shattering rachises. 
The archaeobotanical evidence shows that during the Pre-Pottery Neolithic A (PPNA) cereals exhibited sizes similar to those recorded in domestic species, but their dispersal mechanism was still the same as the one present in morphologically-wild species [i.e. shattering, see @Kirkbride1966; @Kislev1989; @HillmanEtAl2001; @Colledge2001; @WillcoxEtAl2008].
This evidence suggested that wild cereal stands could have been cultivated for as much as a thousand years before non-shattering domestic forms became prevalent in the archaeological record [@TannoWillcox2006; @TannoWillcox2012; @ArranzOtaeguiEtAl2016].
Additional archaeobotanical [@Colledge2001; @WillcoxEtAl2008; @WillcoxEtAl2009; @RiehlEtAl2013; @ArranzOtaeguiEtAl2016; @WeideEtAl2018; @DoucheWillcox2018; @WhitlamEtAl2018] and genetic data [@BadrEtAl2000; @MolinaCanoEtAl2005; @KilianEtAl2007; @OzkanEtAl2011; @IobBotigue2023] in recent years has further challenged the short-gestation model to explain the origins of plant domestication and agriculture in West Asia.

Similarly, the concept of a limited set of eight 'founder crops' [@ZoharyHopf1988] that were the first species cultivated, domesticated and then spread as the basis of Neolithic agricultural systems, is not supported by the latest evidence.
Our previous analyses of the composition of available archaeobotanical datasets shows that these crops were of marginal importance during the Epipalaeolithic period [@ArranzOtaeguiEtAl2018] and that Neolithic subsistence did not rely either solely or primarily on the exploitation of these species [@ArranzOtaeguiRoe2023]. 
Instead, multiple species of grasses, legumes, fruits, nuts, and other plants were exploited over the Late Pleistocene-Early Holocene transition in southwest Asia.

@@

The study of the natural distribution of the progenitors of domesticated crops has been a central part of discussions on the origins of agriculture and plant domestication from the beginning.
@Humboldt1807 acknowledged the importance of the natural distribution of wild species to explain the origin and domestication of crops like spelt and rye.
@Candolle1886 integrated the study of plant ecology and biogeography and influenced @Darwin1859, who later reflected in detail about the geographical distribution of plants and species diversity. 
At that time, there was intense debate about whether there were single or multiple "centres of creation" of species. 
Researchers aimed to evaluate whether plant and animal species emerged in the same locations where they were currently distributed.

After Darwin, this early interest in crop origins evolved into more specific discussions about the "centres of plant domestication".
@Vavilov1926 was among the first to seek to determine the number of regions in which plants had been independently domesticated [@Harris1990].
His main method was 'differential phytogeography':
he classified the variation within a crop and established the regions of maximum diversity, to locate the geographic regions in which crops originated. 
Using this method, Vavilov suggested that there were at least eight centres of origin. 
His work was later criticised by @Harlan1971, who argued that 'centres of origin' and 'centres of diversity' had to be separated. 
For Harlan, a 'centre' was as an "area in which things originate and out of which things are dispersed" [@Harlan1971, p. 468], and he suggested that three main centres of origin of domesticated crops existed.
He further indicated that Vavilov's approach to the question was simplistic and that more data proxies had to be considered (e.g. archaeology, history, geology), an approach more in the tradition of @Candolle1886. 
Indeed, the inclusion of archaeobotany and genetics in the last decades, together with the study of wild relative distributions has been fundamental in characterising the origins of agriculture [@FullerColledge2008]. 
As a result of modern interdisciplinary studies, the number of recognised centres of plant domestication has increased considerably, from the three centres suggested by Harlan in 1971 to the six to eight centres argued for in the 1990s [@Smith1995] and up to as much as 24 potential centres reported in 2009 [@PuruggananFuller2009; see also @Fuller2010].

Biogeographic research into the centres of origin and/or domestication of crops has also long informed broader understanding of the process of agricultural origins.
In *Man Makes Himself*, @Childe1936 correctly located the centre of origin of European agriculture in the 'Fertile Crescent' of West Asia (unlike for example @Pumpelly1908 before him).
This was not based on the region's prehistoric archaeological record, which at that point was barely known.
Instead he was guided to the region by biogeographic work by Vavilov and @PeakeFleure1927;
only later was this prediction validated by archaeological work on the Epipalaeolithic and Neolithic of Palestine [@Boyd2018].
In subsequent decades, the search for more precise origin zones of specific domestic plants relied on the assumption that "the locus of domestication of a wild plant would presumably be within its area of original distribution in the wild state" [@Butzer1971; paraphrasing @Helbaek1959] – and that this "natural habitat" has not changed significantly over the last 12,000 years [@Butzer1971].

Contemporary research on crop origins was pioneered by Harlan and Zohary, who compared the current distribution of the wild progenitors of domesticated plants in southwest Asia [@HarlanZohary1966; @Zohary1969; @Zohary1973; @ZoharyHopf1973] to the rapidly-expanding archaeobotanical record [@Harlan1971; @Harlan1977; see also @ZoharyHopf1988; @HarlanZohary1966].
They both were interested in evaluating which were the wild ancestors of domesticated crops and studying their natural distribution to understand their domestication process [@Zohary1969; @Zohary1973; @ZoharySpiegelRoy1975]. 
Indeed, the natural distribution of the wild relatives of domestic plant species was later used as a criterion to infer 'pre-domestic cultivation' in the archaeological record. 
For example, the presence of seeds of chickpea at Jericho led @Hopf1986 to interpret the remains as cultivars, as the natural distribution of the wild form of chickpea was located further to the north. 
The same rationale was applied to the einkorn remains found at several Pre-Pottery Neolithic sites in the southern Levant [@Hopf1969; @Colledge2001], as the wild progenitors of this species was thought to be restricted to the northern Levantine area [@HeunEtAl1997; @ZoharyEtAl2012]. 
The same idea—presence of plants outside their natural range—has been repeated in the literature more recently by several other authors [@TannoWillcox2006; @WillcoxEtAl2008; @HillmanEtAl2001].

Despite the importance of biogeography in the development and validation of hypotheses regarding the origins of agriculture, there have been few studies of the wild range of specific crop progenitors or other relevant plant species [cf. for domestic animals, e.g. @YeomansEtAl2017].
Observations regarding translocation or range expansion must therefore rely on a relatively rough and ahistoric notion of a species' 'natural distribution' – that is, one based primarily on contemporary or recent-historic occurrences.
Yet we know there has been considerable climatic and environmental change in West Asia since the terminal Pleistocene [@JonesEtAl2019], so it is very unlikely that these ranges were in fact static.
Reconstructions of broader environments have attempted to trace their fluctuations through time, either for the entire region [e.g. @VanZeistBottema1991; Hillman in @MooreEtAl2000] or parts of it [e.g. @Cordova2007], but these are at the level of the vegetation zone rather than individual species.
They also invariably rely on what might be called 'expert interpolation' (where the author composes a map based on his or her own knowledge of the relevant data) rather than an explicit modelling process.
This makes it difficult, if not impossible, for users of such reconstructions to understand exactly how they were derived or what could explain, for example, the significant discrepencies between the predictions of different experts.

@@

Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye). 
Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2001; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].

@@

The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations. 
Since they do accurately predict the test dataset<!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
Another obvious flaw in our methodology is that the time slices used for palaeoclimatic reconstruction are very broad—each covering around two millennia—and therefore potentially unrepresentative of the environment around sites at the specific time at which they were occupied.
The variable quality of the archaeological test dataset, especially in terms of chronology, is also a plausible factor.

@@

"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
Results with regard to specific archaeologically-attested species are not as promising, but this also reflects the incomplete and coarsely temporally-resolved nature of the archaeobotanical verification dataset.
Discrepencies between the modelled distribution, whether on the broad scale (e.g. the restricted geographic range of most species compared to their attestation in the archaeological record), or relating to specific species (i.e. false positives and false negatives), suggest several avenues for future investigation.
